Diffusive Homeostasis in a Self-Organizing Recurrent Neural Network \\ \begin{large} Spatially dependent Interaction as a Determinant of Neural Activity and Plasticity \end{large}

### 2017-02-08 draft.tex
.

\section{Introduction}
\subsection{Homeostasis in the Nervous System}
Among other functional roles of the mammalian central nervous system, it is involved in maintaining crucial properties of the body in a functional regime. Some examples regarding the organism as a whole are e.g. body temperature and blood pressure, both being controlled by hypothalamus \cite[p. 484]{Bear_Exploring_the_Brain}. The term \textit{homeostasis} entailing these regulatory mechanisms was first introduced by Walter C. Cannon in 1932 \cite{Wisdom_of_the_Body_Cannon, Homeostasis_Queenan_2012}. In more recent years, experimental studies have found evidence for processes within the brain that do not function as a regulatory control for certain aspects of the body, but as a homeostatic control of neural activity itself \cite{Turrigiano_1994,Burrone_2003}. Many processes underlying this self-regulatory behavior have been found up till now, and many different biophysical interdependencies have shown to play a role. This diversity makes comparability of these mechanisms a rather difficult task. On a relatively general level though, experiments have revealed two main types of neuronal regulation: synaptic homeostatic plasticity and intrinsic homeostasis \cite{Desai_2003}.

The notion of synaptic homeostatic plasticity subsumes the observation that neurons - being more or less active than a certain set point of mean activity - are regulated by means of their incoming synaptic weights. Theoretical and experimental studies have suggested that this type of regulation is taking place in a multiplicative manner, thereby retaining proportions between individual synaptic strengths \cite{Syn_Plast_Abbott}. Furthermore, one can regard this form of plasticity as an antagonist to other aspects of synaptic plasticity that affect synaptic strengths in an activity dependent, hebbian fashion. Forms of regulation belonging to intrinsic homeostasis alter the neurons' capability of spiking upon external input by changing "intrinsic" properties of the nerve cell. More precisely, a neurons' ability to form an action potential can change by modifying the conductance of ion channels in its cell membrane \cite[p. 156]{Theor_Neur_Dayan}.

Theoretical models of homeostasis are generally based on a form of feedback control. In control theory, this is also known as a \textit{closed loop control}. All these systems include a number of necessary functional compartments, which will be briefly introduced in the following. Concepts and terms were taken from \cite[p. 59-64]{Cybernetik_Systems_Cruse_2006}.
\newpage
\subsubsection{Theory of Control Systems} \label{Control_Sys_Theo_Section}
Generally, a closed loop control must contain some \emph{process} that is to be regulated. Despite its potential internal complexity, its functional role in the context of a feedback system is to simply generate a time-dependent output based on the history of a time-dependent input. This output is what one wishes to control and regulate by means of the feedback loop and thus represents the signal that is further processed. However, it might not be possible to directly quantify this desired output in terms of some internal state of the process. For example, the firing rate of a neuron is an abstract quantity, which can not be \emph{directly} related to some neuronal state variable. Consequently, it might be necessary to transform the actual output of the process into a different signal by a \textit{feedback transducer}. As a necessary next step, the (potentially transformed) output is fed into a \textit{comparator}, which, unsurprisingly, compares the given output of the process to some given target or \textit{reference}, typically by subtraction. Note that this reference can but does note have to be a time-dependent signal. However, the term \textit{homeostasis} generally refers to a fixed reference, while implementations with a time dependent reference can be regarded as a form of \textit{servomechanism}. The difference between reference and the feedback transducer's signal is known as the \textit{error} signal. This error signal is given to a \textit{controller}, which transforms this error signal into an appropriate input signal for the initially introduced process. Additionally, this is also the point in the control loop, where \textit{external disturbances} come into play by being added to the input generated by the controller. We illustrated the described flow of feedback control in Figure \ref{Closed_Loop_Control_Illustration}.
\begin{figure}
\begin{center}
\includegraphics[width=0.7\textwidth]{../../plots/illustrations/closed_loop_control.eps}
\end{center}
\caption{Schematic illustration of a closed loop control as described in Section \ref{Control_Sys_Theo_Section}}
\label{Closed_Loop_Control_Illustration}
\end{figure}
Based on the way the controller generates input for the dynamic process based on the error signal, one can classify control loops by three types:
\newpage
\begin{itemize}
\item A \textit{proportional controller} generates a signal that is proportional to the error signal it receives. The most important property of a proportional controller is that the stationary state of the control loop generally does not imply an error signal equal to zero.
\item An \textit{integrator controller} uses an integration over the past of the error signal to obtain its output. The stationary solution necessarily implies that the error signal is zero.
\item A \textit{differential controller} generates its output by means of the time derivative of the error signal. This type of controller can quickly react to momentary changes but does not compensate for sustained deviations of the error signal from zero.
\end{itemize}
It should be pointed out that these types of control mechanisms can be simultaneously used in a control system, e.g. through a linear combination. Furthermore, the presented terms refer to concepts of linear control theory. Thus, in the study of nonlinear feedback systems they only apply in a limited regime where a first order approximation of the system is sufficiently exact.

\subsubsection{Error Signals and Controllers of Neuronal Homeostasis}
When analyzing or modeling a biological system, a key step characterizing a control loop should be to identify the previously described compartments. While the functional separation illustrated in Figure \ref{Closed_Loop_Control_Illustration} might seem intuitive and "straight forward", it is not obvious which physiological element should be attributed to a certain abstract element.
As a start, we state that the process and its output controlled by homeostasis is the generation of action potentials and the resulting firing rate. As stated earlier, firing rate is an abstract quantity, requiring a feedback transducer to translate the neuron's spiking activity into a "sensible" quantity. In a nerve cell this could be represented by e.g. the relation between the firing of action potentials and the temporary rise in $\mathrm{Ca^{2+}}$ concentration. An example for a homeostatic model making use of intracellular calcium concentration as a measure of spiking activity is the LeMasson-Abbott model \cite{LeMasson_1993}. More complex chains of activity encoding are of course possible, and we will introduce a recent model making additional use of the concentration of nitric oxide in Section \ref{DIff_Hom_Intro_Section} and \ref{NO_Experiments_Section}. Given a representation of activity, its deviation from a reference value will cause other physiological compartments to affect the neuron's ability to generate action potentials. Referring to the concepts introduced in the previous section, one could then ask what type of control best resembles the actual neuronal control. The fact that neurons, whose firing rate was artificially increased or suppressed over a long time, can reliably tune their activity back to the initial state (see e.g. \cite{Burrone_2002}) rules out a purely differential type of control. The return to baseline under different disturbances seems to point towards an integral-like control. However, it is clear that a the ability of a biological system to encode the accumulation of an error signal is limited by the finite amount of available resources. For example, the conductance of an ion channel has natural upper and lower bounds that it can not surpass. We will pick up this issue in Section \ref{Possible_Modifications_Section}.

In any case, modeling homeostasis includes an error estimation, which requires setting some predefined target that the feedback control will attempt to achieve. This, of course, raises the question of how to choose these homeostatic targets. Experimental studies have found that neuronal populations exhibit a strong heterogeneity within their individual activity. In particular, many statistical analyses found evidence for a log-normal like distribution of firing rates \cite{Buzsaki_Fir_Rates_2014,Wohrer_Fir_Rates_2012}. Following these results, one approach of modeling this feature would be to randomly distribute individual targets according to the aforementioned statistics. A problem of this procedure is the absence of any particular deeper physiological motivation or causation. In the light of this conceptual problem, one could, due to a lack of better knowledge, simply define a single global target of activity. However, it has been shown that heterogeneity of neural response onto external input within a population is crucial for efficient stimulus encoding \cite{Marsat_2010}. Allowing for a broad distribution of activity thus has to be regarded an important compartment of models of cortical activity and structure.

>>>>

\section{Introduction}
\subsection{Homeostasis in the Nervous System}
Among other functional roles of the mammalian central nervous system it is involved in maintaining crucial properties of the body in a functional regime. Some examples regarding the organism as a whole are e.g. body temperature and blood pressure, both being controlled by hypothalamus \cite[p. 484]{Bear_Exploring_the_Brain}. The term \textit{homeostasis} entailing these regulatory mechanisms was first introduced by Walter C. Cannon in 1932 \cite{Wisdom_of_the_Body_Cannon, Homeostasis_Queenan_2012}. Experimental studies have found evidence for processes within the brain that do not function as a regulatory control for certain aspects of the body, but as a homeostatic control of neural activity itself \cite{Turrigiano_1994,Burrone_2003}in more recent years. Many processes underlying this self-regulatory behavior have been found up till now and many different biophysical interdependencies have shown to play a role. This diversity makes comparability of these mechanisms a rather difficult task. Though, on a relatively general level experiments have revealed two main types of neuronal regulation: synaptic homeostatic plasticity and intrinsic homeostasis \cite{Desai_2003}.

The notion of synaptic homeostatic plasticity subsumes the observation that neurons - being more or less active than a certain set point of mean activity - are regulated by means of their incoming synaptic weights. Theoretical and experimental studies have suggested that this type of regulation is taking place in a multiplicative manner, thereby retaining proportions between individual synaptic strengths \cite{Syn_Plast_Abbott}. One can regard this form of plasticity as an antagonist to other aspects of synaptic plasticity that affect synaptic strengths in an activity dependent, hebbian fashion. Forms of regulation belonging to intrinsic homeostasis alter the neurons' capability of spiking upon external input by changing "intrinsic" properties of the nerve cell. More precisely, a neurons' ability to form an action potential can change by modifying the conductance of ion channels in its cell membrane \cite[p. 156]{Theor_Neur_Dayan}.

Theoretical models of homeostasis are generally based on a form of feedback control.This is also known as a \textit{closed loop control} in control theory. All these systems include a number of necessary functional compartments, which will be briefly introduced in the following. Concepts and terms were taken from \cite[p. 59-64]{Cybernetik_Systems_Cruse_2006}.
\newpage
\subsubsection{Theory of Control Systems} \label{Control_Sys_Theo_Section}
A closed loop control generally must contain some \emph{process} that is to be regulated. Despite its potential internal complexity, its functional role in the context of a feedback system is to simply generate a time-dependent output based on the history of a time-dependent input. This output is what one wishes to control and regulate by means of the feedback loop and thus represents the signal that is further processed. However, it might not be possible to directly quantify this desired output in terms of some internal state of the process. The firing rate of a neuron for example is an abstract quantity, which can not be \emph{directly} related to some neuronal state variable. Consequently, it might be necessary to transform the actual output of the process into a different signal by a \textit{feedback transducer}.The (potentially transformed) output is fed into a \textit{comparator} in the next step, which - unsurprisingly - compares the given output of the process to some given target or \textit{reference}, typically by subtraction. Note that this reference can but does note have to be a time-dependent signal. The term \textit{homeostasis} however generally refers to a fixed reference, while implementations with a time dependent reference can be regarded as a form of \textit{servomechanism}. The difference between reference and the feedback transducer's signal is known as the \textit{error} signal. This error signal is given to a \textit{controller}, which transforms this error signal into an appropriate input signal for the initially introduced process. This is also the point in the control loop where \textit{external disturbances} come into play by being added to the input generated by the controller. We illustrated the described flow of feedback control in Figure \ref{Closed_Loop_Control_Illustration}.
\begin{figure}
\begin{center}
\includegraphics[width=0.7\textwidth]{../../plots/illustrations/closed_loop_control.eps}
\end{center}
\caption{Schematic illustration of a closed loop control as described in Section \ref{Control_Sys_Theo_Section}}
\label{Closed_Loop_Control_Illustration}
\end{figure}
Based on the way the controller generates input for the dynamic process based on the error signal, one can classify control loops by three types:
\newpage
\begin{itemize}
\item A \textit{proportional controller} generates a signal that is proportional to the error signal it receives. The most important property of a proportional controller is that the stationary state of the control loop generally does not imply an error signal equal to zero.
\item An \textit{integrator controller} uses an integration over the past of the error signal to obtain its output. The stationary solution necessarily implies that the error signal is zero.
\item A \textit{differential controller} generates its output by means of the time derivative of the error signal. This type of controller can quickly react to momentary changes but does not compensate for sustained deviations of the error signal from zero.
\end{itemize}
It should be pointed out that these types of control mechanisms can be simultaneously used in a control system, e.g. through a linear combination. Furthermore, the presented terms refer to concepts of linear control theory. In the study of nonlinear feedback systems they thus only apply in a limited regime where a first order approximation of the system is sufficiently exact.

\subsubsection{Error Signals and Controllers of Neuronal Homeostasis}
A key step in characterizing a control loop in the context of biological systems should be to identify the previously described compartments. While the functional separation illustrated in Figure \ref{Closed_Loop_Control_Illustration} might seem intuitive and "straight forward", it is not obvious which physiological element should be attributed to a certain abstract element.
As a start, we state that the process and its output controlled by homeostasis is the generation of action potentials and the resulting firing rate. Firing rate is - as stated earlier - an abstract quantity, requiring a feedback transducer to translate the neuron's spiking activity into a "sensible" quantity. In a nerve cell this could be represented by e.g. the relation between the firing of action potentials and the temporary rise in $\mathrm{Ca^{2+}}$ concentration. An example for a homeostatic model making use of intracellular calcium concentration as a measure of spiking activity is the LeMasson-Abbott model \cite{LeMasson_1993}. More complex chains of activity encoding are of course possible and we will introduce a recent model making additional use of the concentration of nitric oxide in Section \ref{DIff_Hom_Intro_Section} and \ref{NO_Experiments_Section}. Given a representation of activity, its deviation from a reference value will cause other physiological compartments to affect the neuron's ability to generate action potentials. Referring to the concepts introduced in the previous section one could then ask what type of control best resembles the actual neuronal control. The fact that neurons, whose firing rate was artificially increased or suppressed over a long time, can reliably tune their activity back to the initial state (see e.g. \cite{Burrone_2002}) rules out a purely differential type of control. The return to baseline under different disturbances seems to point towards an integral-like control. It is clear however that a the ability of a biological system to encode the accumulation of an error signal is limited by the finite amount of available resources. For example, the conductance of an ion channel has natural upper and lower bounds that it can not surpass. We will pick up this issue in Section \ref{Possible_Modifications_Section}.

In any case, modeling homeostasis includes an error estimation, which requires setting some predefined target that the feedback control will attempt to achieve. Of course this raises the question of how to choose these homeostatic targets. Experimental studies have found that neuronal populations exhibit a strong heterogeneity within their individual activity. In particular, many statistical analyses found evidence for a log-normal like distribution of firing rates \cite{Buzsaki_Fir_Rates_2014,Wohrer_Fir_Rates_2012}. Following these results, one approach of modeling this feature would be to randomly distribute individual targets according to the aforementioned statistics. A problem of this procedure is the absence of any particular deeper physiological motivation or causation. In the light of this conceptual problem, one could, due to a lack of better knowledge, simply define a single global target of activity. However, it has been shown that heterogeneity of neural response onto external input within a population is crucial for efficient stimulus encoding \cite{Marsat_2010}. Allowing for a broad distribution of activity thus has to be regarded an important compartment of models of cortical activity and structure.
